### 2015-07-12 narrative/data-reduction.tex
first few paragraphs of data refactoring.

The classical workflow where the entire dataset is written to storage for
later analysis is no longer viable simply because the amount of generated data is too large.
In the future, it will be vital to take advantage of a priori
information when 1) writing and reading from the SSIO layer in order to gain
performance and predictability, and to prioritize I/O so that the data that
is most useful for end-user can be written in the time available, and 2)
performing {\em in situ} operations and analysis before storing the information.

In the case of C. S. Chang described earlier, and common with most other
applications scientists with whom we collaborate, information is provided on
which data should be sent over to the SSIO layer so that the best, and most
important information is available for subsequent analysis.
This is generally accomplished using refactoring, which includes re-organization and reductions of the data.

There are many possibilities for refactoring and the best choice will generally
vary from case to case but, once settled, will typically not change from run to run. It
will be important to understand when the time and resources required to
identify and perform the ``best methods'' outweighs any eventual gains. Another
critical research question concerns the quantification and control of the loss
of information resulting from refactoring the data and using a reduced dataset.

A basic issue allied to refactoring is to understand how much information is
actually present in a dataset and whether a refactoring based on a reduced
order representation might prove effective. It is useful to classify scientific
data into two basic categories: structured data (in the sense that it satisfies
a known or relatively simple model), or unstructured data (in which information
content follows no obvious or explicit model).  Although scientific data
generally contains random components (due to finite precision, measurement and
calibration effects etc.), useful scientific data is never purely random.
Broadly speaking, the path from data to knowledge consists of extracting
underlying models or patterns from the datasets, and interpreting the resulting
models.

>>>>

The classical workflow where the entire dataset is written to storage for
later analysis will no longer be viable in exascale, simply because the amount of generated data will 
be too large due to capacity and performance reasons.
In the future, it will be vital to take advantage of \textit{a priori}
information when 1) writing and reading from the SSIO layer in order to gain
higher performance and predictability, and to prioritize the data that
is most useful for end users so that I/O can be finished in the time available, and 2)
performing {\em in situ} operations and analysis before storing the information.
In the fusion use case described earlier which shares commonalities with many other
DOE applications, \textit{a priori} information can indeed be provided by application
scientists regarding which data should be sent over to the SSIO layer so that
minimally, the most science relevant data can be available for subsequent analysis.
This allows the science to be done even when the storage is in busy state - we believe
in a shared multi-user environment, it occurs often \cite{liu_hotstorage}.
In this project, this capability of prioritizing data is accomplished by data refactoring, 
which includes re-organization and reductions of the data.
There are many methodologies to refactor data and the best choice will generally
be application dependent. However, our observation is that, once the choice is settled for an application, 
it will not typically change from run to run.

To refactor data effectively and efficiently, it is important to understand
the critical point when the time and resources required to
identify and perform the ``best methods'' outweighs the intended gains. Another
critical research question concerns the quantification and control of the loss
of information resulting from refactoring the data and using a reduced dataset. 
A basic issue allied to refactoring is to understand how much information is
actually present in a dataset and whether a refactoring based on a reduced
order representation might prove effective. It is useful to classify scientific
data into two basic categories: structured data (in the sense that it satisfies
a known or relatively simple model), or unstructured data (in which information
content follows no obvious or explicit model).  Although scientific data
generally contains random components (due to finite precision, measurement and
calibration effects etc.), useful scientific data is never purely random.
Broadly speaking, the path from data to knowledge consists of extracting
underlying models or patterns from the datasets, and interpreting the resulting
models.
